### 2016-08-31 Chapters/ContinuumSolvationModels.tex
Contributions chapter done
Work on CSM chapter

Developments of the \ac{PCM}, a continuum solvation model, are at the
heart of this thesis.
This Chapter is a brief exposition of the major points of the \acs{PCM}
we have worked upon in the thesis.
Section \ref{sec:IEF} will present some of the mathematical details of the
formulation of the model. I will introduce the concept of \emph{weak}
formulations of \acp{PDE} in Section \ref{sec:weak} and eventually show their
use in the formulation of quantum/classical polarizable Hamiltonians in
Section \ref{sec:variational}.
Section \ref{sec:BEM} will be devoted to the numerical strategies used
in solving the \acs{PCM} equation.

@@

@@

This is a well-known problem in electrostatics~\autocite{Jackson1999, Vanderlinde2005-gf}
and is called a \emph{transmission} problem in the mathematical
literature on \acp{PDE}.~\autocite{Hackbusch1995-uq, Sauter2011-an}
The problem is commonly posed as follows. Given the partition of Euclidean
space $\mathbb{R}^3$ outlined above, find the solution $u(\vect{r})$
to the following set:
\begin{subequations}\label{eq:transmission}
  \begin{align}[left={\empheqlbrace}]
  \Li u &= \nabla^2 u = -4\pi\rhoi \,\, \text{in}\,\, \Omegai \label{eq:internal} \\
  \Le u &= 0 \,\, \text{in}\,\, \Omegae \label{eq:external} \\
  [u] &= \ue - \ui = 0 \,\, \text{on}\,\, \Gamma
  \label{eq:trace-jump} \\
[\partial_L u] &= \partiale u - \partiali u = 0 \,\, \text{on}\,\, \Gamma \label{eq:conormal-jump} \\
|u(\vect{r})| &\leq C \|\vect{r} \|^{-1} \,\,\text{for}\,\,\| \vect{r} \|\rightarrow\infty
\label{eq:radiation}
\end{align}
\end{subequations}
Eqs.~\eqref{eq:trace-jump} and \eqref{eq:conormal-jump}
are the \emph{jump conditions}, expressed in terms of trace operators
for the solution $u$ and its \emph{conormal} derivative.
For notational simplicity, will use the symbols $\partiale$ and $\partiali$
for the latter and only give it in explicit form when needed.
Further mathematical details and technical results on the definition of
function traces and their use in setting up the proper normal and
conormal derivatives can be found in the excellent book by
\citeauthor{Sauter2011-an}.~\autocite{Sauter2011-an}.
The fundamental solutions, or Green's functions, for the elliptic
differential operators $\Li$ and $\Le$ will be denoted by $\Gi$ and
$\Ge$, respectively.
As such the problem is formulated in the so-called \emph{strong} form:
the sought-after solution $u$ has to be at least two times continuously
differentiable.
Moreover, the solution is a function supported over $\mathbb{R}^3$ which
poses challenges to the numerical solution of the problem.
We will take a different approach and reformulate it in terms of
\emph{boundary integral operators}, \ie transform the \acsp{PDE} into a
\ac{BIE}.

The first step is to introduce the relevant boundary integral operators for the
abovementioned transmission problem.
They are three of the four components of the Calder\'on projector.
For any function $v$ in the suitable function space and for $\vect{s}, \vect{s}^\prime \in \Gamma$:
\begin{equation}
\begin{aligned}
  (\bi{S}_\star v)(\vect{s}) &= \int_\Gamma
  G_\star(\vect{s}, \vect{s}^\prime)v(\vect{s}^\prime)\diff \vect{s}^\prime \\
  (\bi{D}_\star v)(\vect{s}) &= \int_\Gamma \partial_{L_\star,\vect{s}}
  G_\star(\vect{s}, \vect{s}^\prime)v(\vect{s}^\prime)\diff \vect{s}^\prime \\
  (\bi{D}^\dagger_\star v)(\vect{s}) &= \int_\Gamma
\partial_{L_\star,\vect{s}^\prime}
  G_\star(\vect{s}, \vect{s}^\prime)v(\vect{s}^\prime)\diff \vect{s}^\prime \\
\end{aligned}
\end{equation}
where $\star$ is a placeholder for the $\mathrm{i}$ or $\mathrm{e}$ subscript.
Our next step is to rewrite the $u$ as the sum of two contributions:
\begin{equation}
  u(\vect{r}) = \esp(\vect{r}) + \xi(\vect{r}) = \int_{\Omegai}
  \frac{\rhoi(\vect{r}^\prime)}{|\vect{r}-\vect{r}^\prime|}
  + \xi(\vect{r})
\end{equation}
the former is the electrostatic potential of $\rhoi$ \emph{in vacuo}
(Newton potential), while the latter is the \emph{reaction} potential.
The reaction potential describes the polarization in the medium.
Moreover, it admits a integral representation as a single-layer
potential:
\begin{equation}
  \xi_\mathrm{i} = \bi{S}_\mathrm{i}\sigma
\end{equation}
where $\sigma$ is the \ac{ASC}, the key quantity in the \acs{PCM}.
As shown by \citeauthor{Cances1998-og}, the \acs{ASC} can be computed
as the \emph{unique} solution to the following integral equation:
\begin{equation}\label{eq:full-IEF}
  \left[ \bi{S}_\mathrm{e}\left(2\pi + \bi{D}^\dagger_\mathrm{i}\right)
  +
  \left(2\pi - \bi{D}_\mathrm{e}\right)\bi{S}_\mathrm{i}
  \right]\sigma =
  -\left[\left(2\pi-\bi{D}_\mathrm{e}\right)
  -\bi{S}_\mathrm{e}\bi{S}_\mathrm{i}^{-1}
  \left(2\pi-\bi{D}_\mathrm{i}\right)
  \right]\esp
\end{equation}
commonly called the \ac{IEF} equation.
The Green's functions for the elliptic differential operators
assume a central role in the boundary integral formulation and make the
\acs{IEF} equation valid for a wide range of external
environments: homogeneous isotropic, ionic and anisotropic
liquids,~\autocite{Cances1998-og}
and systems with interfaces.~\autocite{Corni2002-dr, Frediani2004-er,
Delgado2013-kd, DiRemigio2016-nn}

@@

\section{Detour and Teaser: Weak Formulation of Partial Differential Equations}\label{sec:weak}

From the strong to the weak formulation of \acp{PDE}.\autocite{Ern2004-oo}

\begin{itemize}
  \item Poisson's equation with conductor boundary conditions
        % This is COSMO!
        % This equation makes sense iff \Psi \in \mathcal{C}^2_0(C) (at least)
        % \Psi = 0 \quad \forall \vect{r}\in\partial C
        \begin{equation*}
            \nabla^2\Psi = -4\pi\rho
            % Dirichlet boundary conditions for a conductor
          \quad \Psi \in \mathcolor{ForestGreen}{\mathcal{C}^2_0(C)}
        \end{equation*}
  \item Project onto the space of test functions
        % Project onto the space of test functions
        \begin{equation*}
          \left\{
          \begin{aligned}
            &\text{Seek $\Psi\in\mathcolor{ForestGreen}{V_0}$ such that:}\\
            &(\zeta, \nabla^2\Psi) =
            -4\pi(\zeta, \rho) \quad
            \forall
            \zeta\in\mathcolor{BurntOrange}{\mathcal{C}^\infty_0(C)}
          \end{aligned}
          \right.
        \end{equation*}
  \item Use
    $\eta\nabla^2\gamma = \nabla\cdot(\eta\nabla\gamma) - \nabla\eta\cdot\nabla\gamma$
        % The integral of the first term on the RHS vanishes.
        % It can be shown using Gauss' Theorem and the conductor
        % boundary conditions.
        \begin{equation*}
          \left\{
          \begin{aligned}
            &\text{Seek $\Psi \in \mathcolor{ForestGreen}{H_0^1(C)}$ such that:}\\
            &\mathcolor{Red}{(\nabla\zeta, \nabla\Psi)} =
            \mathcolor{Blue}{-4\pi(\zeta, \rho)} \quad
            \forall \zeta\in \mathcolor{BurntOrange}{H_0^1(C)}
          \end{aligned}
          \right.
        \end{equation*}
\end{itemize}
\begin{equation*}
  \Psi\in \mathcolor{BurntOrange}{H_0^1(C)} = \lbrace f : C \rightarrow
  \mathbb{R} | f \in L^2(C), \nabla f \in L^2(C), f = 0 \forall \vect{s}
  \in\partial C \rbrace
\end{equation*}
\begin{equation}
  \left\{
  \begin{aligned}
    &\text{Seek $\Psi \in V$ such that:}\\
    &\forall \zeta \in V \,\,\,\,\,
    \mathcolor{Red}{a(\zeta, \Psi)} = \mathcolor{Blue}{b(\zeta)}
  \end{aligned}
  \right.
  \label{eq:weak}
\end{equation}

% Hadamard well-posedness definition:
% Problem \eqref{eq:weak} is well-posed if it admits one and only
% one solution and the solution is bounded by the initial data, by
% an a priori estimate (see Ern&Guermond, p.82, Def. 2.1)
\begin{defin}[Well-posedness]\label{def:hadamard}
  The weak problem Eq. \eqref{eq:weak} is \emph{well-posed} if it
  admits one and only one solution and the solution is bounded by
  the initial data by an \emph{a priori} estimate.
\end{defin}
\begin{defin}[Continuity]\label{def:continuity}
  A bilinear form on a normed vector space $V$ is \emph{bounded}, or
  \emph{continuous}, if there is a constant $C$ such that
  $\forall u, v \in V$:
  \[
  a(u, v) \leq C\lVert u\rVert\lVert v\rVert
  \]
\end{defin}
\begin{defin}[Coercivity]\label{def:coercivity}
  A bilinear form on a normed vector space $V$ is \emph{coercive},
  or \emph{elliptic}, if there is a constant $\alpha >0$ such that
  $\forall u \in V$:
  \[
  a(u, u) \geq \alpha\lVert u\rVert^2
  \]
\end{defin}
This implies that no eigenvalue of the linear operator associated
to the bilinear form can be zero, hence that the operator is
invertible!!~\autocite{Ern2004-oo}

% Maybe present connection with Steepest Descent and Conjugate
% Gradient?
% Introduce a suitable vector space, equipped with a norm
% derived from a scalar product.
% The solution belongs to such a vector space, which is
% bigger that in the strong formulation (less constraints on
% the regularity of the solution)
\begin{equation}
  \left\{
  \begin{aligned}
    &\text{Seek $\Psi \in V$ such that:}\\
    &\forall \zeta \in V \,\,\,\,\,
    \mathcolor{Red}{a(\zeta, \Psi)} = \mathcolor{Blue}{b(\zeta)}
  \end{aligned}
  \right.
\end{equation}
\begin{lemma}[Lax--Milgram]
  If the bilinear form $a$ is \emph{continous} and \emph{coercive}
  in $V$, then, for any continuous linear form $b$, Problem~\eqref{eq:weak} is
  \emph{well-posed}.
\end{lemma}
\begin{corollary}[Variational property]
  If the bilinear form is \emph{symmetric} and \emph{positive}
  the unique solution to Problem \eqref{eq:weak} is
  the \emph{unique minimum} on $V$ of the functional:
  \[
  \mathcal{F}(\zeta) = \frac{1}{2}a(\zeta, \zeta) - b(\zeta)
  \]
\end{corollary}

Weak formulation of \acs{BIE}s.\autocite{Hsiao2008-xb}

@@

Approximability, conformity, consistency and asymptotic consistency
of the approximation setting, $\inf$-$\sup$ condition, Céa's Lemma
and generalized $\inf$-$\sup$ condition.~\autocite{Ern2004-oo,
TesiFilippo}
Additional cavity generators
TsLess~\autocite{Pomelli2004-lb}
DefPol~\autocite{Pomelli1998-ob, Pomelli1999-dc}
FIXPVA~\autocite{Su2009-em}
Other discretization schemes
CSC~\autocite{York1999-xy, Scalmani2010-tw}
SWIG~\autocite{Lange2010-jp, Lange2010-qo, Lange2011-eu}
ddCOSMO and ddPCM~\autocite{Cances2013-jg, Lipparini2013-cy, Lipparini2014-to,
Lipparini2014-fq, Stamm2016-fv}

\todo[inline]{CURRENTLY COPIED VERBATIM FROM THE SPHERICAL GREEN'S
FUNCTION PAPER}

The numerical solution of boundary integral equations, such as the
\acs{IEF}-\acs{PCM}
Eqs.~\eqref{eq:full-IEF-2pi}-\eqref{eq:isotropic-IEF}, can be achieved
by the application of the \ac{BEM}.
A discretization of the molecular cavity
boundary is necessary.~\autocite{Pomelli2007-wq}
The discretization of the
molecular surface into finite elements leads to a corresponding
discretization of both the integral operators and the functions in
Eqs.~\eqref{eq:full-IEF-2pi}-\eqref{eq:isotropic-IEF}.

Different strategies are available to perform the
discretization: the \emph{collocation} or the \emph{Galerkin}
approach.~\cite{Hackbusch1995-uq, Ern2004-oo}
The final result is the transformation of
the original integral equation to a system of linear equations whose
dimension relates to the number of finite elements generated by
discretizing the boundary.  For general integral equations and
discretization schemes, a dense system matrix is obtained. Moreover, the
discretization scheme might not, in general, preserve the symmetry
property of the original integral equation.

In this paper we apply the traditional centroid collocation
method~\autocite{Tomasi2005-vm, Pomelli2007-wq} on the cavity boundary
discretization obtained \emph{via} the GePol
algorithm.~\autocite{Pascual-Ahuir1987-uo, Pascual-Ahuir1990-lp,
Pomelli1998-qp, Pomelli2001-sj, Frediani2004-ua, Pomelli2007-wq}
In this algorithm, the cavity is made of interlocking spheres, not all
of them necessarily centered on the atoms. The surface is then
discretized by means of $N_\mathrm{ts}$ small triangles $\Delta_i$ whose
centroid defines the \emph{collocation point}. Once the centroid is
properly defined, functions and operators are discretized by applying a
one-point quadrature rule in the finite element centroid.
The \acs{IEF}-\acs{PCM} equation is thus recast as:
\begin{equation}
  \vect{q}
  =
  \mat{T}^{-1}\mat{R}\vect{v}.
\end{equation}
$\vect{q}$ and $\vect{v}$ are vectors of dimension $N_\mathrm{ts}$
collecting the values of the \acs{ASC} and \ac{MEP}
at the collocation points.
The matrices representing the boundary integral operators are:
\begin{align}
  \mat{T} &=
  \left(2\pi\mat{I} - \mat{D}_\mathrm{e}\mat{A}\right)\mat{S}_\mathrm{i}
  +\mat{S}_\mathrm{e}\left(2\pi\mat{I} +
  \mat{A}\mat{D}_\mathrm{i}^\dagger\right) \\
  \mat{R} &=
  \left(2\pi\mat{I} - \mat{D}_\mathrm{e}\mat{A}\right) -
  \mat{S}_\mathrm{e}\mat{S}^{-1}_\mathrm{i}\left(2\pi\mat{I}-\mat{D}_\mathrm{i}\mat{A}\right)
\end{align}
$\mat{I}$ is the $N_\mathrm{ts}$-dimensional identity matrix while the
other matrices are defined as:
\begin{alignat}{2}
  &\mat{A} = \mathrm{diag}(\cdots\,\, a_i = \mathrm{area}(\Delta_i)\,\, \cdots) \\
  &S_{ii, \mathrm{i}} = k\sqrt{\frac{4\pi}{a_i}}, \quad
  S_{ij, \mathrm{i}} = \frac{1}{|\vect{s}_i-\vect{s}_j|}\quad (i \neq j) \\
  &D_{ii, \mathrm{i}} = -k\sqrt{\frac{\pi}{a_i}}\frac{1}{R_i}, \quad
  D_{ij, \mathrm{i}} =
  \frac{(\vect{s}_i-\vect{s}_j)\cdot\vect{n}_j}{|\vect{s}_i-\vect{s}_j|^3}\quad (i \neq j) \\
  &S_{ii, \mathrm{e}} = \frac{S_{ii, \mathrm{i}}}{C(\vect{s}_i, \vect{s}_i)} +
        G_\mathrm{img}(\vect{s}_i, \vect{s}_i), \quad
  S_{ij, \mathrm{e}} =
  \frac{1}{C(\vect{s}_i, \vect{s}_j)|\vect{s}_i-\vect{s}_j|} +
  G_\mathrm{img}(\vect{s}_i, \vect{s}_j) \quad (i \neq j) \\
  &D_{ii, \mathrm{e}} = \diel(\vect{s}_i)\left[
  \frac{D_{ii, \mathrm{i}}}{C(\vect{s}_i, \vect{s}_i)}
  -S_{ii, \mathrm{i}}
  \frac{\nabla C(\vect{s}_i,
  \vect{s}_i)\cdot\vect{n}_i}{C^2(\vect{s}_i, \vect{s}_i)}
  +
  \nabla G_\mathrm{img}(\vect{s}_i, \vect{s}_i)\cdot\vect{n}_i
  \right] \\
  &D_{ij, \mathrm{e}} =
  \diel(\vect{s}_j)\left[
  \frac{1}{C(\vect{s}_i, \vect{s}_j)}
  \frac{(\vect{s}_i-\vect{s}_j)\cdot\vect{n}_j}{|\vect{s}_i-\vect{s}_j|^3}
  -\frac{\nabla C(\vect{s}_i,
  \vect{s}_j)\cdot\vect{n}_j}{C^2(\vect{s}_i, \vect{s}_j)}
  \frac{1}{|\vect{s}_i-\vect{s}_j|}
  + \nabla G_\mathrm{img}(\vect{s}_i, \vect{s}_j)\cdot\vect{n}_j
  \right]\quad (i \neq j)
\end{alignat}
in terms of the centroids $\vect{s}_i, \vect{s}_j$, the finite elements
areas and their curvatures $R_i$. The factor $k=1.07$ is introduced to achieve a better precision on the diagonal elements~\cite{Tomasi2005}.

\todo[inline]{CURRENTLY COPIED VERBATIM FROM THE WAVELET PAPER}

Boundary integral equations, such as the \acs{IEF}-\acs{PCM}
Eqs.~\eqref{eq:full-IEF-2pi}-\eqref{eq:isotropic-IEF},
can conveniently be solved numerically by the application of the
\ac{BEM}.
Both the integral operators and the functions on which these act are
defined solely on the boundary of the cavity.

The application of the boundary element method requires that the boundary of
the molecular cavity is discretized by a number of suitable finite elements.
The discretization of the boundary leads to a discretization of the integral
operators. This discretization can be carried out by using either the
\emph{collocation} or the \emph{Galerkin}
approach.~\autocite{Hackbusch1995-uq, Ern2004-oo}
In both cases, the integral equation is transformed to a system of
linear equations whose dimension is related to the number of finite
elements used in the discretization of the boundary.
The resulting system matrix is, in general, a dense matrix. The
boundary element method thus suffers from limitations imposed by the
number of matrix elements to be stored and the memory and time
requirements of solving the resulting linear system.

The use of a wavelet basis in the Galerkin approach has been proven
beneficial in this respect.~\autocite{Harbrecht2004-uo, Dahmen2006-pj,
Harbrecht2006-ug}.
The resulting system matrices are quasi-sparse
and can be further reduced to a sparse form by discarding
negligible entries without considerable loss of precision.

The wavelet boundary element method starts by defining
a sequence of hierarchical trial spaces,
spanned by standard finite element ansatz functions:
\begin{equation}
  \{0\} = V_{-1} \subset V_0 \subset V_1 \subset \ldots
  \subset V_J, \qquad V_j = \Span{\phi_{j,k} : k\in \Delta_{j}}.
\end{equation}
Here, $\Delta_j$ is an index set for the
single-scale basis of the space $V_j$. In the wavelet method,
the trial space $V_j$ is split into the direct sum $V_j = V_{j-1}
\oplus W_j$. The resulting complementary space $W_j$ is called the
wavelet space and is not necessarily orthogonal to $V_{j-1}$.
% It suffices the equation $W_j \cap V_{j-1} = \{0\}$,
Recursive splitting of the trial spaces leads to the wavelet
decomposition $V_j = \bigoplus_{l = 0}^{j} W_l$.

>>>>

Developments of the \ac{PCM}, a continuum solvation model, are at the
heart of this thesis.
This Chapter is a brief exposition of the major points of the \acs{PCM}
we have worked upon in the thesis.
Appendix \ref{app:mathematical-results} contains much of the
mathematical details omitted from the exposition in this Chapter.
Section \ref{sec:IEF} will present the formulation of the model.
I will show how a variational approach can facilitate the formulation of
quantum/classical polarizable Hamiltonians in Section
\ref{sec:variational}.
Section \ref{sec:BEM} will be devoted to the numerical strategies used
in solving the \acs{PCM} equation.

@@

This is a well-known problem in electrostatics~\autocite{Jackson1999, Vanderlinde2005-gf}
and is called a \emph{transmission} problem in the mathematical
literature on \acp{PDE}.~\autocite{Hackbusch1995-uq, Sauter2011-an}
Given the partition of Euclidean space $\mathbb{R}^3$ outlined above,
find the solution $u(\vect{r})$ to the following set:
\begin{subequations}\label{eq:transmission}
  \begin{align}[left={\empheqlbrace}]
  \Li u &= \nabla^2 u = -4\pi\rhoi \,\, \text{in}\,\, \Omegai \label{eq:internal} \\
  \Le u &= 0 \,\, \text{in}\,\, \Omegae \label{eq:external} \\
  [u] &= \ue - \ui = 0 \,\, \text{on}\,\, \Gamma
  \label{eq:trace-jump} \\
[\partial_L u] &= \partiale u - \partiali u = 0 \,\, \text{on}\,\, \Gamma \label{eq:conormal-jump} \\
|u(\vect{r})| &\leq C \|\vect{r} \|^{-1} \,\,\text{for}\,\,\| \vect{r} \|\rightarrow\infty
\label{eq:radiation}
\end{align}
\end{subequations}
Eqs.~\eqref{eq:trace-jump} and \eqref{eq:conormal-jump}
are the \emph{jump conditions}, expressed in terms of trace operators
for the solution $u$ and its \emph{conormal} derivative.
For notational simplicity, will use the symbols $\partiale$ and $\partiali$
for the latter and only give it in explicit form when needed.
Further mathematical details and technical results on the definition of
function traces and their use in setting up the proper normal and
conormal derivatives can be found in the excellent book by
\citeauthor{Sauter2011-an}.~\autocite{Sauter2011-an}.
The fundamental solutions, or Green's functions, for the elliptic
differential operators $\Li$ and $\Le$ will be denoted by $\Gi$ and
$\Ge$, respectively.
As such the problem is formulated in the so-called \emph{strong} form:
the sought-after solution $u$ has to be at least two times continuously
differentiable.
Moreover, the solution is a function supported over $\mathbb{R}^3$ which
poses challenges to the numerical solution of the problem.
We will take a different approach and reformulate it in terms of
\emph{boundary integral operators}, \ie transform the \acsp{PDE} into a
\ac{BIE}.

@@

The first step is to introduce the relevant boundary integral operators for the
above mentioned transmission problem.
They are three of the four components of the Calder\'on projector.
For any function $v$ in the suitable function space and for $\vect{s}, \vect{s}^\prime \in \Gamma$:
\begin{subequations}
\begin{align}
  (\bi{S}_\star v)(\vect{s}) &= \int_\Gamma
  G_\star(\vect{s}, \vect{s}^\prime)v(\vect{s}^\prime)\diff
  \vect{s}^\prime \label{eq:S} \\
  (\bi{D}_\star v)(\vect{s}) &= \int_\Gamma \partial_{L_\star,\vect{s}}
  G_\star(\vect{s}, \vect{s}^\prime)v(\vect{s}^\prime)\diff
  \vect{s}^\prime \label{eq:D} \\
  (\bi{D}^\dagger_\star v)(\vect{s}) &= \int_\Gamma
\partial_{L_\star,\vect{s}^\prime}
  G_\star(\vect{s}, \vect{s}^\prime)v(\vect{s}^\prime)\diff
  \vect{s}^\prime \label{eq:D-dagger}
\end{align}
\end{subequations}
where $\star$ is a placeholder for the $\mathrm{i}$ or $\mathrm{e}$ subscript.
Our next step is to rewrite the $u$ as the sum of two contributions:
\begin{equation}
  u(\vect{r}) = \esp(\vect{r}) + \xi(\vect{r}) = \int_{\Omegai}
  \frac{\rhoi(\vect{r}^\prime)}{|\vect{r}-\vect{r}^\prime|}
  + \xi(\vect{r})
\end{equation}
the former is the electrostatic potential of $\rhoi$ \emph{in vacuo}
(Newton potential), while the latter is the \emph{reaction} potential.
The reaction potential describes the polarization in the medium.
Moreover, it admits an integral representation as a single-layer
potential:
\begin{equation}
  \xi_\mathrm{i} = \bi{S}_\mathrm{i}\sigma
\end{equation}
where $\sigma$ is the \ac{ASC}, the key quantity in the \acs{PCM}.
As shown by \citeauthor{Cances1998-og}, the \acs{ASC} can be computed
as the \emph{unique} solution to the following integral equation:
\begin{equation}\label{eq:full-IEF}
  \left[ \bi{S}_\mathrm{e}\left(2\pi + \bi{D}^\dagger_\mathrm{i}\right)
  +
  \left(2\pi - \bi{D}_\mathrm{e}\right)\bi{S}_\mathrm{i}
  \right]\sigma =
  -\left[\left(2\pi-\bi{D}_\mathrm{e}\right)
  -\bi{S}_\mathrm{e}\bi{S}_\mathrm{i}^{-1}
  \left(2\pi-\bi{D}_\mathrm{i}\right)
  \right]\esp
\end{equation}
commonly called the \ac{IEF} equation.
The Green's functions for the elliptic differential operators
assume a central role in the boundary integral formulation and make the
\acs{IEF} equation valid for a wide range of external
environments: homogeneous isotropic, ionic and anisotropic
liquids,~\autocite{Cances1998-og}
and systems with interfaces.~\autocite{Corni2002-dr, Frediani2004-er,
Delgado2013-kd, DiRemigio2016-nn}

@@

@@

The solution of the \acs{IEF}-\acs{PCM} boundary integral equation can
only be achieved analytically for simple cavity geometries, such as
single spheres and ellipsoids.
When more general molecular cavities are employed one has to resort to a
numerical technique: the \ac{BEM}. This is implemented in four steps:
\begin{enumerate*}[label={\alph*)},font={\color{PMS1797}}]
    \item define the molecular cavity
    \item mesh the molecular surface \label{item:mesh}
    \item discretize the integral equation \label{item:discretize} and
    \item solve the linear systems by a suitable numerical technique.
\end{enumerate*}
Many different approaches to these four steps have been presented, both
in the field of the \acs{BEM} and in the context of the \acs{PCM}.

The molecular cavity is usually defined as a set of atom-centered
interlocking spheres. The radii are chosen from a suitable set, usually
van der Waals radii.~\autocite{Bondi1964-dt, Mantina2009-hb}
If one imagines to model a solvent molecule as a rigid sphere,
three types of molecular surfaces can be defined:
\begin{enumerate*}[label={\alph*)},font={\color{PMS1797}}]
  \item
    the union of the interlocking spheres surfaces is the \ac{vdWS},
    Figure \ref{fig:vdWS}
  \item the locus of points defined by the center of the solvent
    spherical probe while rolling over the \acs{vdWS} is the
    \ac{SAS}, Figure \ref{fig:SAS}
  \item the locus of points defined by the contact point of the solvent
    spherical probe while rolling over the \acs{vdWS} is the \ac{SES}, Figure \ref{fig:SES}.
\end{enumerate*}
\begin{figure}[tb]
 \centering
 \begin{subfigure}{.45\textwidth}
   \centering
   \includegraphics[width=.85\textwidth]{vdWS.png}
   \caption{\acl{vdWS} (\acs{vdWS})}
   \label{fig:vdWS}
 \end{subfigure}
 ~
 \begin{subfigure}{.45\textwidth}
   \centering
   \includegraphics[width=.85\textwidth]{SAS.png}
   \caption{\acl{SAS} (\acs{SAS})}
   \label{fig:SAS}
 \end{subfigure}

 \begin{subfigure}{.45\textwidth}
   \centering
   \includegraphics[width=.85\textwidth]{SES.png}
   \caption{\acl{SES} (\acs{SES})}
   \label{fig:SES}
 \end{subfigure}
 \caption{Molecular surfaces from interlocking sets of atom-centered spheres.
 Reproduced from \noparcite[ref.][]{Harbrecht2011-dk}.}
\end{figure}
Alternative definitions based on the isodensity surface of the molecular solute
have been put forth.~\autocite{Foresman1996-wv}

Steps \ref{item:mesh} and \ref{item:discretize} are closely intertwined,
in the technical \acs{BEM} literature they go under the name of
\emph{finite element discretization}.
Let us define a mesh of the domain $\Gamma$ as an indexed collection of
intervals with non-zero measure $\lbrace I_i\rbrace_{i=1}^{N_\mathrm{mesh}}$, where
$N_\mathrm{mesh}$ is the size of the mesh.
Then, in a rather informal sense, a finite element is the
mathematical entity formed by the vector space of functions that are
piecewise regular on any given interval in the mesh.~\autocite{Ern2004-oo}
Basically, given an element in the mesh, we "attach" a basis set of
piecewise polynomials of a prescribed degree and use this basis to
provide a representation of quantities defined on $\Gamma$ within each
mesh interval.
Mesh generation for the molecular surfaces mentioned above is usually
achieved by modified triangulation procedures. In this thesis we used
the GePol algorithm~\autocite{Pascual-Ahuir1987-uo, Pascual-Ahuir1990-lp,
Pomelli1998-qp, Pomelli2001-sj, Frediani2004-ua, Pomelli2007-wq}
and the mesh generators of Harbrecht and Randrianarivony especially
tailored for the wavelet \acs{BEM}.~\autocite{Harbrecht2009-no,
Harbrecht2011-dk}

The finite element discretization is usually carried out either by the
\emph{collocation} or by the \emph{Galerkin} approach.\footnote{
Indeed, it is possible to show that discretization by collocation
corresponds to a Galerkin method with a very specific (and asymmetric)
choice of \emph{trial} and \emph{test}
spaces.~\autocite{Hackbusch1995-uq, Ern2004-oo}}
This step transforms the original integral equation into a system of
linear equations whose dimension relates to the underlying mesh size.
For general integral equations and finite element discretizations,
the \emph{stiffness matrix} obtained is dense and proper linear algebra
techniques need to be employed in order to avoid a detrimental impact on
computational performance.
In addition, finite element discretization might destroy intrinsic
symmetries of the original integral
operators. These can be recovered by appropriate \emph{a posteriori}
symmetrization procedures introducing, however, a degree of
arbitrariness in the \acs{BEM} procedure.~\autocite{Lange2011-eu}

In the GePol procedure, the molecular surface is meshed by means of
$N_\mathrm{ts}$ spherical polygons $\Pi_i$ whose centroids define the
collocation points. One-point quadrature rules are then applied at the
centroid to obtain the matrix equation:
\begin{equation}
 \mat{R}^{-1}\mat{T} \vect{q}
  =
  \vect{v}.
\end{equation}
$\vect{q}$ and $\vect{v}$ are vectors of dimension $N_\mathrm{ts}$
collecting the values of the \acs{ASC} and \ac{MEP}
at the collocation points. $\mat{R}^{-1}\mat{T}$ is the stiffness
matrix.
The matrices representing the boundary integral operators are:
\begin{align}
  \mat{T} &=
  \left(2\pi\mat{I} - \mat{D}_\mathrm{e}\mat{A}\right)\mat{S}_\mathrm{i}
  +\mat{S}_\mathrm{e}\left(2\pi\mat{I} +
  \mat{A}\mat{D}_\mathrm{i}^\dagger\right) \\
  \mat{R} &=
  \left(2\pi\mat{I} - \mat{D}_\mathrm{e}\mat{A}\right) -
  \mat{S}_\mathrm{e}\mat{S}^{-1}_\mathrm{i}\left(2\pi\mat{I}-\mat{D}_\mathrm{i}\mat{A}\right)
\end{align}
$\mat{I}$ is the $N_\mathrm{ts}$-dimensional identity matrix while the
other matrices are defined as:
\begin{subequations}
  \begin{align}
    A_{ij} &= a_i\delta_{ij}, \quad a_i = \mathrm{area}(\Pi_i) \\
    S_{ij, \mathrm{i}} &=
    k\sqrt{\frac{4\pi}{a_i}}\delta_{ij} +
    \frac{1}{|\vect{s}_i-\vect{s}_j|}(1-\delta_{ij}) \\
    D_{ii, \mathrm{i}} &=
    -k\sqrt{\frac{\pi}{a_i}}\frac{1}{R_i}\delta_{ij}
    + \frac{(\vect{s}_i-\vect{s}_j)\cdot\vect{n}_j}{|\vect{s}_i-\vect{s}_j|^3}(1-\delta_{ij})
  \end{align}
\end{subequations}
in terms of the centroids $\vect{s}_i, \vect{s}_j$, the finite elements
areas and their curvatures $R_i$.
The factor $k=1.07$ is introduced to achieve a better precision on the
diagonal elements.~\autocite{Tomasi2005-vm}
Expressions for $\mat{S}_\mathrm{e}$, $\mat{D}_\mathrm{e}$ and
$\mat{D}^\dagger_\mathrm{e}$ depend on the specific outer environment
and can be found in the literature.~\autocite{Corni2002-dr, Frediani2004-er,
Tomasi2005-vm, Delgado2013-kd, DiRemigio2016-nn}

The wavelet Galerkin \acs{BEM} was also used in this thesis. This method
preserves the fundamental symmetries of the boundary integral operators
and achieves sparsity in the stiffness matrix.
This sparsity can be leveraged to implement efficient algorithms with
linear space and time complexity. The approach is thus superior
to the traditional collocation method.

Given a hierarchical sequence of trial spaces
$\lbrace V_l\rbrace_{l=1}^J$, one can decompose the $l$-th element in
the hierarchy into the direct sum $V_l = V_{l-1} \oplus W_l$.
$W_l$ is called the \emph{wavelet space} and is the, not necessarily
orthogonal, complement to $V_{l-1}$.
Recursively applying this splitting to the trial spaces hierarchy
generates the wavelet decomposition, where all complementary spaces are
spanned by the wavelet basis.
The fundamental insight in the wavelet \acs{BEM} is to use the wavelet
basis for the discretization step and employ the compression technique
described by \citeauthor{Dahmen2006-pj} to build up the sparse system
matrix.~\autocite{Harbrecht2004-uo, Harbrecht2004-ug, Dahmen2006-pj}
Matrix compression is carried out in two rounds, \emph{a priori} and \emph{a
posteriori}, resulting in a finger-like sparsity pattern for the
stiffness matrix, Figure \ref{fig:finger-structure}.
The number of relevant matrix entries scales proportional to $O(N_J)$
with $N_J$ the number of degrees of freedom for a refinement of the
geometry up to level $J$.

\begin{figure}[tb]
   \includegraphics[width=\textwidth]{a_2_l_0_5_fs.png}
   \caption[Sparsity pattern achieved by the wavelet Galerkin \acs{BEM}.]{
   Sparsity pattern achieved by the wavelet Galerkin \acs{BEM} for a
   model problem (Laplace equation on 6 patches).
   Reproduced from \noparcite[ref.][]{Bugeanu2015-tp}.
   }
   \label{fig:finger-structure}
\end{figure}
